Reinforcement-Learning Windturbine Controller

### 2019-12-03 paper/paper.tex
Added death condition handling, fixed some bugs, wrote PER section

\copyrighterklaerung{Hier soll jeder Autor die von ihm eingeholten
Zustimmungen der Copyright-Besitzer angeben bzw. die in Web Press
Rooms angegebenen generellen Konditionen seiner Text- und
Bild"ubernahmen zitieren.}
\acknowledgments{Die Danksagung...}

@@

Similar to other reinforcement learning algorithms, we will again not compute the full gradient but use stochastic gradient ascend with batch learning. To rephrase gradient ascend to gradient descend, they multiply the policy loss with $-1$. In fact, in the original \ac{DDPG} algorithm, they used Adam \cite{kingmaAdamMethodStochastic2017} optimization instead of stochastic gradient descend.

@@

Unfortunately, we are not able to provide a scientific comparison between different methods, as our learning proved too instable to derive sane metrics from it. Thus, this section will be structured as an assortment of problems and their solutions

@@

At the start of our experimentation, we took off with a well-known problem and a proven implementation. We used the OpenAI-Pendulum task, in which a pendulum in gravity needs to be held upright by applying a force to it. OpenAI Gyms generally serve as benchmarks for reinforcement learning algorithms and most foundational papers and articles evaluate their solutions based on how well they perform in OpenAI Gyms. Starting with a reference-implementation in pytorch by \cite{DeepDeterministicPolicy}, we experimented with the pendulum environment. We observed that the algorithm got stuck in a local optimum, where it rotates the pendulum instead of holding it upright, for long periods of the training process. Only with high noise, we could get it to hold a pendulum upright. We also tried to run the HalfCheetah task, where a two-dimensional dog-like being should be brought to running and moving its joints, but the results were more a crawling and shaking towards the goal. However, we hoped that the implementation might yield results on our windturbine environment already and exchanged the OpenAI gym for qblade.

@@

In \cite{mnihPlayingAtariDeep2013} they use an epsilon-greedy strategy, where with a certain small probability epsilon, a random action is taken. As we have a continous action space, we can add a little bit of noise towards every action: \(a' = a + n \). We first tried a gaussian noise: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \). We didn't see any improvements over our old behavior, so we switched to an Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930} as recommended in the \ac{DDPG} paper \cite[formula (7)]{lillicrapCONTINUOUSCONTROLDEEP2015}. Additionally to the \ac{DDPG} paper, we let the sigma parameter decay over time, so that the noise gets less when training has proceeded.

@@

The TD3-paper \cite{fujimotoAddressingFunctionApproximation2018} utilizes a random exploration phase, in which a completely random policy \cite{mozerDiscoveringStructureReactive1990} initially explores the environment, before the actual agent switches in. We implemented the random walk at first through gaussian noise and then through an Ornstein-Uhlenbeck process. The later version creates less vibration through less excessive changes in control parameters. As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process. We set the parameters to that controller by experimentation as we did not need the actual control parameters of the controller. However it could be imagined to use an established controller here. The PID controller was easily able to keep the windturbine in a sensible operation range, and without noise it resulted in a smooth and constant operation. Though our PID controller is certainly inferior to industry-grade controllers, we will view it as an expert controller which is already achieving a certain degree of quality.

@@

We tackled this problem by at first reducing the number and size of layers from the example implementation to one fully connected layer of 64 neurons in the critic and one fully connected layer of 32 neurons in the actor. The reference-implementation of \cite{DeepDeterministicPolicy} utilized 3 levels in the critic with 1024, 512 and 300 fully connected neurons and in the actor 2 levels of 512 and 128 fully connected neurons. Reducing complexity partially tackled the problem of zero gradients, we could then observe small gradients which however vanished over time.

@@

More effectively, we added data normalization. We normalized the data so states, actions and reward always stayed between 1 and -1. This finally fixed our problem of zero gradients and we could observe normal learning. We recommend this technique as a default technique for every reinforcement learning algorithm. As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others. As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were not considered due to their small absolute values. At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values. As sometimes after random exploration, extreme values were observed, we started normalising on the replay data of a previous run. We could also imagine manually setting normalization bounds on low-dimensional problems, as sometimes human knowledge about the possible state space is present.

@@

\cite{schaulPrioritizedExperienceReplay2016} proposed a method to sample experiences from the replay buffer according to how much they benefit training, and not just uniformly. In their paper, they showed how especially cliffwalk problems, but effectively all problems, benefit from using prioritized experience replay. In their case, they use it for \ac{DQN}, but we will apply it to \ac{DDPG}, as our windturbine problem is also such a problem in which rewards are sparse.

As knowing how much a sample benefits training is difficult, they propose to approximate this importance by the temporal difference error $\tau = y - Q(s, a)$. This difference, which in Q learning is comparable to the loss of the approximator, tells us how far off the prediction was in this step. So, alongside with the experiences, a sampling probability for each item $p_i$ is stored in the replay buffer. To be able to efficiently sample from it, we implemented it as proposed in their paper in the form of a sum-tree.

When replaying, the sampling probability of a sample is set to $P(j) = \frac{p_j^{\alpha}}{\sum_i{p_i^{\alpha}}}$, $\alpha$ being a hyperparameter allowing to fade between uniform sampling ($\alpha=0$) and only priority sampling ($\alpha=1$). New samples are always stored with maximum priority, and to correct for oversampling bias, a correction term is applied to the temporal difference error: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$. It is simply multiplied to the real TD-error. $\beta$ is a parameter describing how strongly to correct oversampling, where $\beta=1$ is full correction and $\beta=0$ no correction at all.

We decided to split policy and critic updates, as the oversampling effects could also happen to policy updates, and still sample uniformly for policy updates.

@@

\subsection{Huber loss}
We could already greatly improve our results by using huber-loss \cite{huber1964} for the critic instead of mean squared error loss as described in DDPG paper. However we still saw some loss explosions.

@@

\subsection{Batch normalization}
As in deep learning, batch normalization is helpful for increasing lr and thus to skip over local maxima, we tried to add batch normalization between the layers. The authors suspected problems with the noise in RL, however it looked good.

\section{Automatized hyperparameter search}

To test all the above things, we tried simulations with different hyperparameters.

@@

It must be stated that we did not achieve a single run which stayed within sensible operational range for longer than 30K steps. Even our handcrafted PID controller for which we took the first set of parameters that somehow works widely outperforms our learnt policies, an industry controller is even better. This result in itself is notable, because it needed a whole set of additions to plain \ac{DDPG} to get first sensible results on our windturbine environment. So apparently, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance, or we are having errors in our programme. As the final product we developped is close to \cite{fujimotoAddressingFunctionApproximation2018} and has improvements from a wide variety of papers, we conclude that tackling wind turbine control with \ac{RL} will need more research. Though we were not able to outperform industry controllers, the speed in which the area of \ac{RL} is advancing makes us hopeful to achieve good results in the near future. With our work on integrating qblade, we provided an open-source, accessible framework to develop future reinforcement learning controllers for wind turbines.

>>>>

\copyrighterklaerung{Hier soll jeder Autor die von ihm eingeholten
Zustimmungen der Copyright-Besitzer angeben bzw. die in Web Press
Rooms angegebenen generellen Konditionen seiner Text- und
Bild"ubernahmen zitieren.}
\acknowledgments{At first thanks to my advisors Mattew Lennie and Dmitrij Shlezinger for their support. Also, I guess, thanks David Silver - 7 of my most important papers were from you. Thanks to my laptop for almost dying, but then surviving. And big thanks to the mighty god of windturbines for letting me crush $\num{12345}$ windturbines for science - yes I counted.}

@@

Similar to other reinforcement learning algorithms, we will again not compute the full gradient but use stochastic gradient ascend with batch learning. To rephrase gradient ascend to gradient descend, they multiply the policy loss with $-1$. In fact, in the original \ac{DDPG} algorithm, they used Adam \cite{kingmaAdamMethodStochastic2017} optimization instead of stochastic gradient descend. As activations, they used relu \cite{agarapDeepLearningUsing2019}. Their network architecture for the actor consisted of two fully connected layers with 400 and 300 units and instead of relu they used tanh activations in the last layer to bound actions. For the critic, they had a version which learnt upon pixel outputs and used convolutional layers, while we only look at the low-dimensional version. For this, they also used a two layer variant with 400 and 300 units. State inputs were fed through the whole network, while actions were concatenated in onto the second layer, skipping the first.

Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}

@@

For everyone who is interested in a recap of our design process, in this section we will provide such. We will again present our algorithm in a later section. To understand our design decisions, it might be interesting to read this passage, but you can also skip this passage and just read the rest.

@@

At the start of our experimentation, we took off with a well-known problem and a proven implementation. We used the OpenAI-Pendulum task, in which a pendulum in gravity needs to be held upright by applying a force to it. OpenAI Gyms generally serve as benchmarks for reinforcement learning algorithms and most foundational papers and articles evaluate their solutions based on how well they perform in OpenAI Gyms. We managed to find a solution with our DDPG implementation. We hoped that the implementation might yield results on our windturbine environment as well and exchanged the OpenAI gym for qblade.

@@

We first tried a gaussian noise instead of the recommended \ac{OU-Noise} \cite{uhlenbeckTheoryBrownianMotion1930} from \ac{DDPG}: \(n \leftarrow \mathcal{N}(\mu, \sigma^2) \). We didn't see any improvements over our old behavior, so we switched to \ac{OU-Noise}. Additionally to the \ac{DDPG} paper, we let the sigma parameter decay over time, so that the noise gets less when training has proceeded. We hope to only need this noise in the first stage of the training and then, later, see better results. Our decay is parameterized by a start sigma, end sigma and end step and then held constant at the end sigma.

@@

The TD3-paper \cite{fujimotoAddressingFunctionApproximation2018} utilizes a random exploration phase, in which a completely random policy \cite{mozerDiscoveringStructureReactive1990} initially explores the environment, before the actual agent switches in. We implemented the random walk at first through gaussian noise and then through an Ornstein-Uhlenbeck process. The later version creates less vibration through less excessive changes in control parameters. As we found that both random policies rarely explore a sensible operational range, we later added an expert PID controller, which implements torque control, and combined that with our Ornstein-Uhlenbeck process. We set the parameters to that controller by experimentation and came up with the following equation, observing rotational speed $rot$ from the environment: $(torque, pitch) = (\num{2e7}*rot-\num{12e6}, 128*rot-103)$. These actions were clipped to our action bounds. So effectively, we designed a P controller, and only using this, we could already achieve relatively stable control. As further work, it could be imagined to use an established controller here, but we did not need this for our purposes. The PID controller was easily able to keep the windturbine in a sensible operation range, and without adding noise it resulted in a smooth and constant operation.

@@

We tackled this problem by at first reducing the number and size of layers from the example implementation to one fully connected layer of 64 neurons in the critic and one fully connected layer of 32 neurons in the actor. Reducing complexity partially tackled the problem of zero gradients, we could then observe small gradients which however vanished over time.

@@

More effectively, we added data normalization. We normalized the data so states, actions and reward always stayed between 1 and -1. This is a normal technique in machine learning, and we are irritated by the fact that this is not part of the default algorithm. \cite{vanhasseltLearningValuesMany2016} propose an integration mechanism into \ac{RL}, but we decided to implement a more straight-forward variant. At first, we used observed states and rewards from the random exploration phase to calculate maximum/minimum values. As sometimes after random exploration, no extreme values were observed, we started normalising on the min/max of replay data of a previous run. We could also imagine manually setting normalization bounds on low-dimensional problems, as sometimes human knowledge about the possible state space is usually present.
This finally fixed our problem of zero gradients and we could observe normal learning. We recommend this technique as a default technique for every reinforcement learning algorithm. As in our simulation, where our state reflects measurements from very different parts of the simulation, some values in the state-array were 8 orders of magnitude different from others. As most of the high-magnitude values are vibrations, the net could not create a link between these and the state-action, while low-magnitude but insightful observations like rotational speed were not considered due to their small absolute values.

@@

\cite{schaulPrioritizedExperienceReplay2016} proposed a method to sample experiences from the replay buffer according to how much they benefit training, and not just uniformly. This method is called \ac{PER} According to how much they benefit means that we add a priority to the experiences, which we somehow derive from the performance of the algorithm on that samples. In their paper, they showed how especially cliffwalk problems, but effectively all problems, benefit from using prioritized experience replay. We argue that our windturbine is similar to such a cliffwalk, as random exploration usually only shortly passes sensible operation range and then quickly destroys the turbine. Seeing this short high reward might not be enough for the controller to learn that being there is good. In the original case, they use it on \ac{DQN}, but we will apply it to \ac{DDPG}. There hasn't been much work on how to use \ac{PER} in conjunction with \ac{DDPG}, in fact we only found two papers: \cite{houImprovingDDPGPrioritized} and \cite{zhaExperienceReplayOptimization2019}. We will at first present the general prioritized experience replay algorithm, then we will discuss the papers and then present how we incorporated \ac{PER}.

As knowing how much a sample benefits training is difficult, \cite{schaulPrioritizedExperienceReplay2016} propose to approximate this importance by the \ac{TD-Error} $\delta = y - Q(s, a)$. This difference, which in Q learning is comparable to the loss of the approximator, tells us how far off the prediction was in this step. So, alongside with the experiences, a sampling priority for each item $p_i$ is stored in the replay buffer. In theory, we would have to calculate this \ac{TD-Error} every time we update our critic. This would mean to feed our entire replay buffer through the critic on every update step. Clearly, this is a huge performance hit, which is why in the algorithm, they approximated by only updating the error on samples that are anyways being used for training. We calculate the \ac{TD-Error} anyways for the critic update, so we can store this without performance penalty. New samples are always stored with maximum priority to make sure they are sampled at least once.

When replaying, the sampling probability of a sample is set to $P(j) = \frac{p_j^{\alpha}}{\sum_i{p_i^{\alpha}}}$, $\alpha$ being a hyperparameter allowing to fade between uniform sampling ($\alpha=0$) and pure priority based sampling ($\alpha=1$). Because we introduced oversampling bias by sampling some experiences more often than others, a correction term is applied to the \ac{TD-Error}: $w_j = \frac{(N*P(j))^{-\beta}}{\max_i{w_i}}$. It is simply multiplied to the real \ac{TD-Error}. $\beta$ is a parameter describing how strongly to correct oversampling, where $\beta=1$ is full correction and $\beta=0$ no correction at all. This importance sampling is a common technique in reinforcement learning and we have addressed it before in our background part.

In addition to sampling directly based on $\abs{\delta}$, \cite{schaulPrioritizedExperienceReplay2016} propose a variant where they sample according to the rank of an experience: $p_i = \frac{1}{\text{rank}(i)}$ where rank is defined as the position in the replay memory when sorted according to $\abs{\delta}$. This method is more robust to outliers and slightly increases performance, but because of the extra computational effort of having to calculate the rank every time, we decided to go for the first method. Also, \cite{zhaExperienceReplayOptimization2019} measured worse performance of the rank based version in contrast to even plain \ac{DDPG}.

The importance sampling is also the reason why we deem it not so trivial to introduce \ac{PER} to \ac{DDPG}, as there is no specified way how to integrate this into policy updates. But first we will describe two other efforts. \cite{houImprovingDDPGPrioritized} implemented it straightforward on the Q-Learning part of \ac{DDPG} and ignored any effects on policy updates. They measured improvements tackling the OpenAI Pendulum task in contrast to vanilla \ac{DDPG}. However, they only evaluated this pendulum task, which we already suspect not to be representative of our windturbine problem. We prefer to rely on the work of \cite{zhaExperienceReplayOptimization2019} - the team has formulated experience replay as a learning problem and they trained another prediction network on which replay to choose. They evaluated their learnt experience replay against sampling based on rank or priority and found their algorithm to outperform both, while rank-based sampling performed far worse than even vanilla \ac{DDPG} and priority based sampling slightly better than vanilla \ac{DDPG}. However, the team completely disregarded importance sampling in their work. Also, we deem it as unnecessary to introduce yet another learning task, especially because we are calculating on weak hardware.

Because neither of the works seemed solid enough for us to implement it, but because the learning improvements in Q-learning are very promising, we decided to create our own mixture of \ac{PER} and \ac{DDPG}. We decided to implement the Q-Learning part straightforward as in \cite{schaulPrioritizedExperienceReplay2016} and including importance sampling. We used the reformulated \ac{TD-Error} of \ac{DDPG} with the target actor and critic. To avoid oversampling bias in the policy, we only applied \ac{PER}-sampling to the training part of the critic and trained the actor on uniformly sampled data. It can be argued, that as the policy $\mu$ is anyways being trained on the state density function of another policy $\rho^{\beta}$, it can deal with that oversampling bias. However we doubt adding more bias is beneficial to the training, so we decided to sample uniformly for the policy.

Because we saw performance problems storing the priorities alongside the experiences and directly sampling from a 100000-item weighted array, we followed the advice of \ac{PER} to use a sumtree datastructure, a binary weighted tree in which each node holds the sum of weights of the children. We implemented sampling as a tree walk starting at the root node. At every node, we normalized the priorities of the two child nodes to probabilities, did a random draw and recursively repeated until we reached a leaf. This recursive sampling technique retains the sample probabilities of the leaf nodes but is cheaper in complexity than to normalize the whole array each timestep and do a random draw over the whole probability distribution. All our operations on the replay buffer are in $O(log(n))$ complexity now.

@@

\subsection{Huber loss}
We could already greatly improve our results by using huber-loss \cite{huber1964} for the critic instead of mean squared error loss as described in DDPG paper. In fact, the authors of \ac{DQN} have caused some confusion on this in the \ac{RL}-community. In their journal paper \cite{mnihHumanlevelControlDeep2015}, they write \"We also found it helpful to clip the error term from the update [...] to be between -1 and 1\", which could be interpreted as loss clipping. However they actually switched to absolute error instead of square error after those boundaries, which is the definition for huber loss.

Although loss explosions gotten less frequent and smaller in magnitude with this update, we still saw some.

@@

\subsection{Batch normalization}
As in deep learning, batch normalization is helpful for increasing lr and thus to skip over local maxima, we tried to add batch normalization between the layers. We did get a performance boost when using this, however we ended up with NaN values after a while of running it, especially when experiencing q loss explosions. As these unexpected nan-values made training more complicated, we decided to not use batch normalization.

\subsection{Regarding death conditions}

In \ac{DQN}, death conditions are accounted for by changing the Q target in the terminal step to be $y = r$ instead of $y = r + \gamma \max_a' Q(s', a')$. This way, behavior after a death will not be accounted for by the Q function. As we are seeing many death conditions, we thought it could be useful to include this into our DDPG algorithm. So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions.

\chapter{Algorithm}

After our experimentation phase, we will present our algorithm, on which we perform our evaluation. At first, we describe how we encapsulated the qblade environment. At second, we will present our replay buffer, our network structure, our exploration strategy and our update algorithm.

\section{QBlade}

As mentioned before, we set the yaw angle to always 0 degrees, as in reality controlling the orientation of the nacelle is trivial and in our simulations, wind will always come from the front. The generator torque on our NREL 5MW turbine has a sensible range of \num{0} to ca \num{4e6} Nm, we used a range of \num{0} to \num{8e6} as action space, as higher torque can be simulated by qblade and this way made it easier for the controller, as it can apply high torques to slow down quickly rotating blades.
Because changing torque from zero to maximum in one timestep is not realistic for a real turbine, we restrict it to move a maximum of \num{3e5} Nm per one timestep of 0.1 seconds. If our agent chooses a value outside of that area, we set it to the closest value inside of that area.
Blades can be pitched between 0 and 90 degrees, 0 being not pitched out at all and operating at maximum efficiency and 90 resulting in no aerodynamic torque from the rotor whatsoever. A sensible pitch motor can only turn the blades at a limited speed, we assume a limit of 5 degrees per second and again set controller inputs to a sensible value inside that. The simulation theoretically accepts a full pitch change in one timestep, however as inertia on the blade is so high for such a change, the blades break instantly and the rest of the simulation needs to be reset.

QBlade simulates blade damage, also high rotational speeds cause blades to fall off. We however observed problems with our controller when reaching these extreme limits of the simulation In the course of our experiments, we decided to reset the simulation at an rotational speed of 3 rad/s, as our turbine normally operates at ca 0.8 rad/s. Also, high generator torque inputs cause the simulation to rotate the rotor backwards with a negative energy yield, effectively creating a multimillion-dollar leafblower. As this is neither a realistic scenario, we also reset the simulation at -0.5 rad/s. We tried to clip at 0, but this resulted in so frequent deaths that learning actions at this border was not possible.

When resetting the simulation, we let it run for 100 timesteps with zero action and state, as the elastic structure needs a bit of time to move from simulation default to elastic equilibrium. In this time, simulation results do not make sense yet. Also, every 100000 timesteps or after seeing NaN values we reload the project file to maximize simulation stability.

We hid all observations except for power and rotational speed from the algorithm and also merged all pitch controls into a single variable.

\section{Core algorithm}

As DDPG is capable of learning from a replay of experiences, we implemented a replay buffer. To enable \ac{PER}, we needed to sample according to priorities proportional to a metric, in our case \ac{TD-Error}. We implemented the replay buffer as a sum tree datastructure, a binary weighted tree in which each node holds the sum of the weights of the nodes below. The root node thus always held the sum of all metrics. Also, we did not store probabilities right away but kept the unnormalized experiences $p_i = (\abs{\sigma} + \epsilon)^{\alpha}$ with an epsilon of $\num{1e-6}$ and $\alpha = 0.5$.

We implemented our actor model according to the \ac{DDPG} paper, but with less neurons per layer. Instead of 400 and 300 neurons, we use 32 and 16 neurons. Our critic is also according to the \ac{DDPG} paper but with less neurons. Instead of 400 and 300 we use 64 and 32 neurons. All states, rewards and actions seen by the network are normalized to be between $[-1, 1]$.

We start our simulation with an expert exploration. We designed a P-controller which can keep the turbine at a constant rotational speed and added \ac{OU-Noise} to it to explore sensible operation range. For 3 epochs, we let our expert policy with noise control the turbine before we hand over to our \ac{DDPG}-Agent. During the handover, the policy is pretrained for 10k steps, lr 1e-4, adam optimizer, mse loss to directly predict actions from the expert controller. This way we started off early with better actions. Then, critic and actor training is done on the experiences from expert exploration. After this training, all priorities in the replay buffer are set to their maximum (so they all will be seen) and normal training starts. Each epoch lasts for 2000 steps and at the beginning of it, the simulation is reset to default state. We trained for 150 epochs.

Critic updates are done similar to \ac{DQN} updates with \ac{PER}. We use importance sampling to correct for \ac{PER} overestimation bias and regard death conditions.

@@

It must be stated that we did not achieve a single run which stayed within sensible operational range for longer than 30K steps. Even our handcrafted PID controller for which we took the first set of parameters that somehow works widely outperforms our learnt policies, an industry controller is even better. This result in itself is notable, because it needed a whole set of additions to plain \ac{DDPG} to get first sensible results on our windturbine environment. So either, our windturbine simulation is more complex to learn than the OpenAI gym examples, which are usually used as benchmarks for \ac{RL} performance, or we are having errors in our programme. As the final product we developped has improvements from a wide variety of papers but still does not produce stable results, we conclude that tackling wind turbine control with \ac{RL} will need more research. Though we were not able to outperform industry controllers, the speed in which the area of \ac{RL} is advancing makes us hopeful to achieve good results in the near future. With our work on integrating qblade, we provided an open-source, accessible framework to develop future reinforcement learning controllers for wind turbines.

To demonstrate the instability, we ran 5 training runs with the same hyperparameters and measured return, epoch return, killcount and time to q explosion (where we define q explosion as q loss > 1e4)
